Fractional Types

### 2016-12-18 typespp.tex

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not categorically equivalent and hence $\simeq$ cannot be categorical
equivalence.  There are however other notions of groupoid equivalence, such as
Morita equivalence~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
identify the two groupoids above.  Although the notion of equivalence we use in
the next section appears novel, it is related ``in spirit'' to Morita
equivalence and we find it useful to give a small example. We will verify that
the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is the trivial
groupoid with only one distinct object and one trivial automorphism. To avoid
clutter we assume the groupoids are strict, e.g., that the groupoid
$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
elements). The definition of \emph{Morita equivalence} in this case reduces to
to finding a \emph{Morita morphism} from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
from the unit groupoid to the product groupoid, say $G$, with an additional
condition that the commutative diagram in figure~\ref{fig:morita} is a
\emph{pullback}. The intuition is that the codomain consists of three objects
with associated morphisms, each of them counting as a ``third.'' The functor
from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor. To summarize the
equivalence between the two groupoids requires a \emph{local} map that chooses a
target and a \emph{global} condition ensuring that the choice ultimately
annihilates. The challenge we are facing is to turn this idea into an operational
semantics that somehow combines the local and global aspects of the
equivalence. We will be inspired by the folklore result that ``substitution into a
dependent type is a pullback''~\cite{subpullback} as explained in the next section.

\begin{figure}[h]
  \centering
  \begin{tikzcd}
    Q
    \arrow[drr, bend left, "{\xi_1}"]
    \arrow[ddr, bend right, "{(s,t)}"]
    \arrow[dr, dotted] & & \\
    & \set{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
    & G_1 \arrow[d, "{(s,t)}"] \\
    & \set{\ast} \times \set{\ast} \arrow[r, "{\psi_0 \times \psi_0}"]
    & G_0 \times G_0
  \end{tikzcd}
  \caption{
    \label{fig:morita}
    Pullback diagram for a Morita morphism,
    $\order{\idiso} \rightarrow \order{\permfive} \circledast
    \iorder{\permfive}$
  }
\end{figure}

@@

@@

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We then allow indices of values $k$ to be range over a collection
of types. In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' sharing the same $k$.

>>>>

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
\[
\order{p} \circledast \iorder{p} \simeq \order{\idiso}
\]
The left hand side is the categorical product of the iteration groupoid and
symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
groupoid is 1. The right hand side is the iteration groupoid for the identity
1-combinator which also has cardinality 1. The groupoids on either side are
however not categorically equivalent and hence $\simeq$ cannot be categorical
equivalence.  There are however other notions of groupoid equivalence, such as
Morita equivalence~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
identify the two groupoids above.  Although the notion of equivalence we use in
the next section appears novel, it is related ``in spirit'' to Morita
equivalence and we find it useful to give a small example. We will verify that
the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is equivalent
to the trivial
groupoid with only one distinct object and one trivial automorphism. To avoid
clutter we assume the groupoids are strict, e.g., that the groupoid
$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
elements). The definition of \emph{Morita equivalence} in this case reduces to
to finding a \emph{Morita morphism} from $\order{\idiso}$ to
$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
from the unit groupoid to the product groupoid, say $G$, with an additional
condition that the commutative diagram in figure~\ref{fig:morita} is a
\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps,
and $\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids
defined previously.
The intuition is that the codomain consists of three objects
with associated morphisms, each of them counting as a ``third.'' The functor
from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
include an arbitrary choice of ``which third'' to target and the universality
condition of the pullback ensures that the choice is ultimately insignificant as
it leads to the same result as every other possible functor. To summarize the
equivalence between the two groupoids requires a \emph{local} map that chooses a
target and a \emph{global} condition ensuring that the choice ultimately
annihilates. The challenge we are facing is to turn this idea into an operational
semantics that somehow combines the local and global aspects of the
equivalence. We will be inspired by the folklore result that ``substitution into a
dependent type is a pullback''~\cite{subpullback} (but which turns out to have
been fully explained in~\cite{practicalfoundations}) as explained in the next
section.

\begin{figure}[h]
  \centering
  \begin{tikzcd}
    Q
    \arrow[drr, bend left, "{\xi_1}"]
    \arrow[ddr, bend right, "{(s,t)}"]
    \arrow[dr, dotted] & & \\
    & \order{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
    & G \arrow[d, "{(s,t)}"] \\
    & \ot \times \ot \arrow[r, "{\psi_0 \times \psi_0}"]
    & \mathbb{3} \times \mathbb{3}
  \end{tikzcd}
  \caption{
    \label{fig:morita}
    Pullback diagram for a Morita morphism,
    $\order{\idiso} \rightarrow \order{\permfive} \circledast
    \iorder{\permfive}$
  }
\end{figure}

@@

%%%%%
\subsection{Dependent Pointed Types}\label{sec:deppointedtypes}

To understand the challenge in designing an operational semantics for $\eta$ and
$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
manipulation on types:
\[
x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
\]
and by the coherence conditions of compact closed categories, we expect this
circuit, when run in either direction, to be equivalent to the
identity. Consider now the case of a combinator $c$ whose order is greater than
or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
either end. The problem becomes apparent when one observes that the uses of
$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
right-to-left execution must behave differently depending on the incoming
value. But the ``local'' information available to either of these combinators
does not include the ``global'' information about the incoming value and there
is no way for them to make consistent guesses locally. It is possible to modify
the operational semantics to use a mechanism to express a computational effect
which enables spatially separated parts of the program to communicate in a way
that is reminiscent of entanglement in quantum mechanics. Possibilities to
realize this communication are global reference cells, backtracking, or logical
variables with unification. Instead, we choose a more abstract approach that
encodes the required dependency in dataflow constraints using dependent types.

\begin{definition}[Dependent $\Pi^/$]
\label{def:pifracd}
The syntax of dependent $\Pi^/$ differs from Def.~\ref{def:pifrac} as follows:
\[\begin{array}{lrcl}
(\textrm{Fractional Types}) & \mathbb{T_\bullet} &::=&
  [\order{c},c^k] \alt
  [\iorder{c},1/c^k] \alt
  [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
  \forall k.\mathbb{T_\bullet} \\
 (\textrm{/-combinators}) & \rho_\bullet &:&
  \mathbb{T}_{\bullet 1} \leftrightblackspoon \mathbb{T}_{\bullet 2} ~::=
  [\textit{see Fig.~\ref{pifracd:comb}}]
\end{array}\]
\end{definition}

@@

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We then allow universal quantification over indices $k$
associated to values, over a collection of types.
In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' sharing the same $k$.
It is important to remember that there is no loss in generality
in going to pointed types, as all fractional types are inhabited (see
remark in Sec.~\ref{sec:symmetryg}.
